---
title: "Reproducible Research Course Project 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```


### Notes

Following scripts use functions from tidyverse library which includes dplyr, ggplot2 and other useful packages. 






### Loading and Preprocessing the Data

```{r load}

data <- read.csv("activity.csv")

```
#### Converting the type of date variable from factor to date

```{r date}

data$date <-  as.Date(data$date,  "%Y-%m-%d")


```

### Part I: What is mean total number of steps taken per day?


#### 1. Total steps taken by date computed using dplyr functions group_by and summarize 


```{r total steps}

data2 <- data %>% group_by(date) %>% summarise(Totalsteps = sum(steps)) %>% print()


```

#### 2. Histogram of the total number of steps taken each day


```{r histogram}

hist(data2$Totalsteps, main = "Histogram of Total Steps Taken Each Day", xlab = "Total Steps")


```

#### 3. The mean of total steps taken with missing values removed


```{r mean}

mean(data2$Totalsteps, na.rm = T)


```

#### 4. The median of total steps taken with missing values removed

```{r median}

median(data2$Totalsteps, na.rm = T)


```


### Part II: What is the average daily activity pattern?


#### 1. Time series plot of 5 minute intervals (x axis) and average number of steps taken averaged across all days (y axis)

``` {r plot}


data3 <- data %>% group_by(interval) %>% summarise(Meansteps = mean(steps, na.rm = T))

plot(data3, type="l", main= "Average Steps Taken By Hour")



```


#### 2. The 5 minute interval containing highest amount of average steps taken

``` {r max}


data3[which.max(data3$Meansteps),]

```
### Part III: Imputing missing values

#### 1. Total number of missing values in the dataset

``` {r NA}

sum(is.na(data))
```

#### 2. Filling the missing values with means of 5 minute intervals and creating a new dataset

``` {r filling NAs}

data4 <- data %>% group_by(interval) %>% mutate_at(vars(steps), ~replace(., is.na(.), mean(steps, na.rm = T))) %>% ungroup()

```

#### 3. Histogram of the total number of steps with NAs filled

``` {r no NA histogram}

data5 <- data4 %>% group_by(date) %>% summarise(Totalsteps = sum(steps))

hist(data5$Totalsteps, main = "Histogram of Total Steps Taken Each Day (NAs filled)", xlab = "Total Steps")
```

#### 4. Mean and median total number of steps taken per day with NAs filled

``` {r mean and median}
mean(data5$Totalsteps, na.rm = T)
median(data5$Totalsteps, na.rm = T)
```

#### Comments: As we can see imputing missing values has almost no effect on mean and median values.


### Part IV: Are there differences in activity patterns between weekdays and weekends?


#### 1. Creating a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

``` {r week}

data4 <- data4 %>% mutate(Week = weekdays(date))
data4$Week <- as.factor(data4$Week)
data4$Week <- fct_collapse(data4$Week, weekday = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"), 
                           weekend =  c("Saturday", "Sunday"))

```

#### 2. Panel plot showing time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)


``` {r panel plot}

data6 <- data4 %>% group_by(interval, Week) %>% summarise(Meansteps = mean(steps))  

ggplot(data6, aes(x =interval, y = Meansteps)) + labs(y = "Average Number of Steps", y = "Year") +
  geom_line(color = "blue", stat="identity") + facet_grid(rows = vars(data6$Week)) + guides(fill=FALSE)


```

#### Comments: We can see that on weekdays, presumably due to work, number of steps spike around 08.30 then fall down while on weekends it shows a more even distribution.
